---
title: "Main"
output: html_document
---

# 1. Imports, data loading and initial data inspection

## 1.1 Imports

```{r include=FALSE}
library(ggplot2)
library(tidyr)
library(viridisLite)
library(viridis)
library(dplyr)
library(sf)
library(leaflet)
library(geojsonio)
library(htmltools)
library(tidyverse)
library(scales)
library(psych)
library(manipulateWidget)
```

## 1.2 Data Loading and Initial Data Inspection

```{r}
df_wb = read.csv("wellbeing.csv")
df_mr = read.csv("material_flow.csv")
```

```{r}
head(df_wb)
```

```{r}
colnames(df_wb)
```

```{r}
head(df_mr)
```

```{r}
colnames(df_mr)
```

```{r}
find_unit <- function(value) {
  if (value >= 1e9) {
    return ("Gt")
  } else if (value >= 1e6) {
    return ("Mt")
  } else if (value >= 1e3) {
    return ("kt")
  } else {
    return ("t")
  }
}
```

```{r}
fix_years_col <- function(years) {
  if(length(years) == 0) {
    return(c())
  }
  return (paste0("X", as.character(years)))
}
```

```{r}
plot_flow_category <- function(data, flow, category, years, ylabel=FALSE, debug=FALSE) {
  years_columns <- fix_years_col(years)
  years_filter <- c("Country", years_columns)  
  data_filtered <- data %>%
    filter(Flow.name == flow, Category == category) %>%
    select(all_of(years_filter)) %>%
    drop_na()
  
  maximum_year <- max(as.numeric(years))
  max_col <- fix_years_col(as.character(maximum_year))
  
  data_filtered <- data_filtered %>%
    filter(.data[[max_col]] > 0) %>%
    arrange(desc(.data[[max_col]])) %>%
    head()
  
  if (debug) {
    print(data_filtered)
    return()
  }
  
  melted_df <- data_filtered %>%
    pivot_longer(
      cols = -Country,
      names_to = "Year",
      values_to = "Value"
    )
  
  country_order <- melted_df %>%
    filter(Year==max_col) %>%
    arrange(desc(Value)) %>%
    pull(Country)
  
  melted_df <- melted_df %>%
    mutate(Country=factor(Country, levels=country_order))
  
  print(melted_df)
  ggplot(melted_df, aes(x=Country, y=Value, fill=Year)) +
      geom_bar(stat="identity", position="dodge") +
      scale_fill_viridis_d() +
      theme_minimal() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
      xlab("Country") +
      ylab(if (!isFALSE(ylabel)) ylabel else {
        max_value <- max(melted_df$Value, na.rm = TRUE)
        paste0(flow, " (", find_unit(max_value), ")")
      }) +
      ggtitle(paste0(flow, " by Country for Selected Years")) +
      guides(fill = guide_legend(title = "Year"))
}
```

## 2. Filtering and cleaning data

### 2.1 Filtering functions

```{r}
filter_minerals <- function(data, flow_name, category, year = 2023) {
  year <- fix_years_col(year)
  cols_to_select <- c("Country", year)
  data_filtered <- data %>%
    filter(Category==category, Flow.name==flow_name) %>%
    select(cols_to_select) %>%
    dplyr::rename(c(Minerals=year)) %>%
    drop_na()
  
  return(data_filtered)
}
```

```{r}
filter_minerals_generic <- function(data, flow_name="", category="", years=c()) {
  years <- fix_years_col(years)
  cols_to_select <- c("Country", "Category", "Flow name", years)
  data <- data %>%
      filter(Category==category, Flow.name==flow_name) %>%
      select(-one_of(c("Flow.code", "Flow.unit")))
  if(length(years) == 1) {
    year_str = years[1]
    data <- data %>%
          select(Country, !!year_str) %>%
          rename(c(Minerals=!!year_str))
  } else if(length(years) > 1) {
    data <- data %>%
          select(Country, Category, Flow.name, all_of(years)) %>%
          pivot_longer(
            cols=all_of(years),
            names_to="Year",
            values_to="Value"
          )
  } else {
    data <- data %>%
          pivot_longer(
            cols=-one_of(c("Country", "Category", "Flow.name")),
            names_to="Year",
            values_to="Value"
          )
  }
  
  data <- data %>% drop_na()
  return(data)
}
```

```{r}
filter_wellbeing <- function(data, measure, domain, years = c(2023), age = "total", sex = "total", edu = "total") {
  age <- tolower(age)
  sex <- tolower(sex)
  edu <- tolower(edu)
  
  age <- paste0(toupper(substr(age, 1, 1)), substr(age, 2, nchar(age)))
  sex <- paste0(toupper(substr(sex, 1, 1)), substr(sex, 2, nchar(sex)))
  edu <- paste0(toupper(substr(edu, 1, 1)), substr(edu, 2, nchar(edu)))
  
  if (age == "Mid") {
    age <- "Middle-aged"
  }
  
  if (edu != "Total") {
    edu <- paste(edu, "education")
  }
  
  data_filtered <- data %>%
    filter(
      Measure == measure,
      Domain == domain,
      Age == age,
      Sex == sex,
      Education.level == edu
    )
  
  print(data_filtered)
  all_years = unique(data$TIME_PERIOD)
  
  if(length(years) == 0) {
    data_filtered <- data_filtered %>%
                   filter(TIME_PERIOD %in% all_years)
  } else {
    data_filtered <- data_filtered %>%
                   filter(TIME_PERIOD %in% years)
  }
  
  return(data_filtered)
}
```

### 2.2 Dropping unused/duplicate columns and rows

```{r}
remove_rows_from_dataset <- function(rows, df) {
  df <- df %>% filter(!Country %in% rows)
  return(df)
}
```

```{r}
remove_columns_from_dataset <- function(columns, df) {
  columns <- intersect(columns, colnames(df))
  df <- df %>% select(-all_of(columns))
  return(df)
}
```

```{r}
rows_to_remove <- c("World", "Asia + Pacific", "North America", "EECCA",
                    "West Asia", "Africa", "Europe", "Latin America + Caribbean")
df_mr <- remove_rows_from_dataset(rows_to_remove, df_mr)
head(df_mr)
```

```{r}
columns_to_remove <- c("STRUCTURE", "STRUCTURE_ID", "STRUCTURE_NAME", "ACTION", "MEASURE", "UNIT_MEASURE", "Unit of measure", "AGE", "SEX", "EDUCATION_LEV", "DOMAIN", "Time period", "Observation value", "OBS_STATUS", "UNIT_MULT", "Unit multiplier", "Units", "DECIMALS", "Decimals", "Base period", "BASE_PER")
df_wb <- remove_columns_from_dataset(columns_to_remove, df_wb)
head(df_wb)
```

## 2.3 Check for missing values

```{r}
missing_per_columns_minerals <- colSums(is.na(df_mr))
```

Missing values per column in minerals dataset:

```{r}
print(missing_per_columns_minerals[missing_per_columns_minerals > 0])
```

```{r}
missing_per_column_wellbeing <- colSums(is.na(df_wb))
```

Missing values per column in wellbeing dataset:

```{r}
print(missing_per_column_wellbeing[missing_per_column_wellbeing > 0])
```

## 2.4 Replacing names for correct map plot assignations

```{r}
unique(df_mr["Country"])
```

```{r}
replace_country_names <- function(df, name_dict, column="name") {
  matches <- match(df[[column]], names(name_dict))
  df[[column]][!is.na(matches)] <- name_dict[matches[!is.na(matches)]]
  return(df)
}
```

```{r}
country_replacements <- c(
  "Russian Federation" = "Russia",
  "Czechia" = "Czech Republic",
  "Viet Nam" = "Vietnam"
)
```

```{r}

df_mr <- replace_country_names(df_mr, country_replacements, column = "Country")
df_wb <- replace_country_names(df_wb, country_replacements, column = "Reference.area")
```

## 2.5 Unique values for columns of interest

```{r}
categories <- unique(df_mr["Category"])
categories
```

```{r}
flows <- unique(df_mr["Flow.name"])
flows
```

```{r}
measures <- unique(df_wb["Measure"])
measures
```

```{r}
domains <- unique(df_wb["Domain"])
domains
```

## 2.6 Map plotting helper functions

```{r}
plot_map_static <- function(data, year) {
  # Load shapefile
  shapefile_path <- "map/ne_110m_admin_0_countries.shp"
  world <- st_read(shapefile_path, quiet = TRUE)
  
  # Merge the data with the shapefile
  world_data <- world %>%
    left_join(data, by = c("NAME" = "Country"))
  
  # Create the plot
  ggplot(data = world_data) +
    geom_sf(aes(fill = year), color = "black", size = 0.2) +
    scale_fill_distiller(palette = "RdBu", direction = 1, na.value = "gray90") +
    theme_minimal() +
    labs(title = "World Map with Values per Country", fill = year) +
    theme(axis.text = element_blank(),
          axis.ticks = element_blank(),
          panel.grid = element_blank())
}
```

```{r}

plot_map <- function(data, column = "Minerals", title = "", wh = "100x100") {
  year <- names(data)[2]
  geo_json_path <- "map/folium/world-countries.json"
  
  geo_json_data <- geojson_read(geo_json_path, what = "sp")
  
  dims <- strsplit(wh, "x")[[1]]
  width <- paste0(dims[1], "%")
  height <- paste0(dims[2], "%")
  
  geo_json_data@data <- geo_json_data@data %>% left_join(data, by = c(name="Country"))
  
  pal <- colorNumeric("YlOrRd", domain = geo_json_data@data[[column]], na.color = "transparent")
  m <- leaflet(geo_json_data) %>%
       addTiles() %>%
       addPolygons(
         fillColor = ~pal(get(column)),
         weight = 0.3,
         opacity = 1,
         color = "black",
         fillOpacity = 0.7,
         highlight = highlightOptions(
           weight = 2, color = "black", fillOpacity = 0.8, bringToFront = TRUE
         ),
         label = ~paste0(name, ": ", get(column)),
         labelOptions = labelOptions(direction = "auto")
       ) %>%
       addLegend(pal = pal, values = ~get(column), title = "Value per Country")
  if (title != "") {
    title_html <- tags$div(
      HTML(paste0("<h1 style='position:absolute;z-index:1000;left:40vw'>", title, "</h1>"))
    )
    m <- htmlwidgets::prependContent(m, title_html)
  }
  
  m
}
```

```{r}

plot_with_slider <- function(data, flow_name, category) {
  columns_to_ignore <- c("Flow.code", "Flow.unit")
  df_filtered <- data %>%
                  select(-one_of(columns_to_ignore)) %>%
                  filter(Flow.name==flow_name,Category==category)
  
  df_melted <- df_filtered %>%
               pivot_longer(cols = -c("Country", "Category", "Flow.name"), names_to = "Year", values_to = "Value")

  geo_json_path <- "map/folium/world-countries.json"
  geo_json <- st_read(geo_json_path, quiet=TRUE)
  
  geo_countries <- geo_json$name
  df_melted <- df_melted %>% filter(Country %in% geo_countries)
  year_columns <- grep("^X\\d{4}$", colnames(df_filtered), value = TRUE)
  max_year_numeric <- max(as.numeric(sub("X", "", year_columns)))
  min_year_numeric <- min(as.numeric(sub("X", "", year_columns)))
  
  manipulateWidget(
    {
      df_year <- df_melted %>% 
                 filter(Year==paste0("X", as.character(year)))
      max_val <- max(df_year$Value, na.rm=TRUE)
  
      geo_data <- geo_json %>%
                  left_join(df_year, by=c(name="Country"))
      
      pal <- colorNumeric("YlOrRd", domain = c(0, max_val), na.color = "transparent")
  
      leaflet(geo_data) %>%
        addProviderTiles(providers$CartoDB.Positron) %>%
        addPolygons(
          fillColor = ~pal(Value),
          weight = 1,
          color = "orange",
          fillOpacity = ~Value/max_val,
          label = ~paste(name, ": ", round(Value, 2)),
          highlight = highlightOptions(
            weight = 2, color = "#666", fillOpacity = 0.8, bringToFront = TRUE
          )
        ) %>%
        addLegend(pal = pal, values = geo_data$Value, title = paste(category, flow_name, year, " "))
    },
    year = mwSlider(min_year_numeric,
                  max_year_numeric,
                  step = 1,
                  value = 2020,
                  label = "Year")
  )
}
```

```{r}

plot_wellbeing_with_slider <- function(data, title="") {
  geo_json_path <- "map/folium/world-countries.json"
  geo_json <- st_read(geo_json_path, quiet=TRUE)
  
  geo_countries <- geo_json$name
  df <- data %>% filter(Reference.area %in% geo_countries)
  max_year_numeric <- max(df$TIME_PERIOD)
  min_year_numeric <- min(df$TIME_PERIOD)
  
  manipulateWidget(
    {
      df_year <- df %>% 
                 filter(TIME_PERIOD==year) %>%
                 rename(c(Country="Reference.area", Year="TIME_PERIOD", Value="OBS_VALUE"))
      max_val <- max(df_year$Value, na.rm=TRUE)
  
      geo_data <- geo_json %>%
                  left_join(df_year, by=c(name="Country"))
      
      pal <- colorNumeric("YlOrRd", domain = c(0, max_val), na.color = "transparent")
  
      leaflet(geo_data) %>%
        addProviderTiles(providers$CartoDB.Positron) %>%
        addPolygons(
          fillColor = ~pal(Value),
          weight = 1,
          color = "orange",
          fillOpacity = ~Value/max_val,
          label = ~paste(name, ": ", round(Value, 2)),
          highlight = highlightOptions(
            weight = 2, color = "#666", fillOpacity = 0.8, bringToFront = TRUE
          )
        ) %>%
        addLegend(pal = pal, values = geo_data$Value, title = title)
    },
    year = mwSlider(min_year_numeric,
                  max_year_numeric,
                  step = 1,
                  value = 2020,
                  label = "Year")
  )
}
```

# 3. Extended Analysis

## 3.1 Global variables stored

```{r}
years = c(1970, 1997, 2024)
df_mr_transposed <- filter_minerals_generic(df_mr)
```

## 3.2 Visualization

### 3.2.1 Univariate Analysis

```{r}
annual_earnings <- filter_wellbeing(df_wb, measure="Average annual gross earnings", domain="Work and job quality")
annual_earnings <- annual_earnings %>% arrange(desc(OBS_VALUE))

annual_earnings

ggplot(annual_earnings, aes(x=reorder(Reference.area, -OBS_VALUE), y=OBS_VALUE, fill=Reference.area)) +
    geom_col(show.legend = FALSE) +
    scale_colour_viridis(
      alpha = 1,
      begin = 0,
      end = 1,
      direction = 1,
      discrete=FALSE,
      option="D",
      aesthetics="color"
    ) + 
    labs(
      title="Average Annual Gross Earnings (2023)",
      x="Country",
      y="Earnings (in USD)"
    ) +
    theme_minimal(base_size=14) +
    theme(
      axis.text.x = element_text(angle=45, hjust=1),
      panel.grid.major.y = element_line(linetype="dashed", color="gray80")
    )
```

```{r}
df_mr_dist <- df_mr %>%
  filter(Flow.name=="Domestic Extraction",Category == "Fossil fuels") %>%
  select("X1970", "X2023") %>%
  drop_na()

summary_stats <- psych::describe(df_mr_dist)
stats_df <- as.data.frame(summary_stats)
stats_df$Year <- c("X1970", "X2023")

stats_df <- stats_df %>%
            pivot_longer(
              cols=-Year,
              names_to = "Statistics",
              values_to = "Values"
            )
print(stats_df)
ggplot(stats_df, aes(x = Statistics, y = Values, fill = Year)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_y_log10(labels=scales::comma) +
  labs(
    title = "Descriptive Statistics for Minerals dataset (1970 vs 2023)",
    x = "Statistic",
    y = "Value (log scale)"
  ) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "grey")
```

```{r}
minerals_melted <- df_mr %>%
  pivot_longer(
    cols = -c(Country, Category, Flow.name, Flow.code, Flow.unit),
    names_to = "Year",
    values_to = "Value"
  )

avg_value_per_year <- minerals_melted %>%
                      group_by(Year) %>%
                      summarise(Value=mean(Value, na.rm=TRUE)) %>%
                      ungroup()
avg_value_per_year <- avg_value_per_year %>% mutate(Value = Value / 1e6)

ggplot(avg_value_per_year, aes(x=Year, y=Value, group=1)) +
      geom_line(color="#0072B2") +
      scale_x_discrete(breaks = avg_value_per_year$Year[seq(1, nrow(avg_value_per_year), by = 2)]) +
      labs(
        title="Average Material Flow Value Over Years",
        x="Year",
        y="Average Domestic Extraction (Million tonnes)"
      ) +
      theme_minimal(base_size=14) +
      theme(axis.text.x = element_text(angle=45, hjust=1))
```

Observation: An analysis of material flow data reveals notable declines in the years 2009 and 2020, coinciding with significant global economic downturns. These patterns underscore the sensitivity of material flows to global economic conditions, highlighting how major crises can lead to substantial reductions in resource extraction and consumption.

### 3.2.2 Multivariate Analysis

```{r}
plot_flow_category(df_mr, flow="Domestic Extraction", category="Fossil fuels", year=years)
```

Observation: In the above plot, the major players in the industry of domestic extraction of fossil fuels are displayed, along with their evolution on 3 equidistant year landmarks.

```{r}
plot_flow_category(df_mr, flow="Exports", category = "Metal ores", years=years)
```

Observation: Above it is displayed a plot about the major countries that export metal ores, on the same 3 year landmarks

```{r}
plot_with_slider(df_mr, "Domestic Extraction", "Fossil fuels")
```

```{r}
plot_with_slider(df_mr, flow_name = "Imports", category = "Metal ores")

```

```{r}
plot_with_slider(df_mr, flow_name = "Exports", category = "Metal ores")
```

```{r}
plot_with_slider(df_mr, flow_name = "Exports", category = "Fossil fuels")
```

```{r}
df_employment_rate = filter_wellbeing(df_wb, measure="Employment rate", domain="Work and job quality", years=c())
plot_wellbeing_with_slider(df_employment_rate)
```

```{r}

```

```{r}
plot_with_slider(df_mr, flow_name="Material Footprint (RMC)", category = "Biomass")
```

```{r}
life_satisfaction <- filter_wellbeing(df_wb, measure="Life satisfaction", domain="Subjective well-being", years=c(2018)) %>%
    rename(c(Country="Reference.area", Year="TIME_PERIOD", Value="OBS_VALUE"))
domestic_extraction_fossil_fuels <- filter_minerals_generic(df_mr, flow_name="Domestic Extraction", category="Fossil fuels", year=c(2018))

life_satisfaction

merged1 <- merge(life_satisfaction, domestic_extraction_fossil_fuels, by="Country")

min_val <- min(merged1$Minerals, na.rm=TRUE)
max_val <- max(merged1$Minerals, na.rm=TRUE)

merged1$Minerals_Scaled <- 10 * (merged1$Minerals - min_val) / (max_val - min_val)

ggplot(merged1, aes(x=Value, y=Minerals_Scaled, label=Country)) +
    geom_point() +
    geom_text(vjust=0.5, hjust=0.5, size=3) +
    labs(
      x="Life satisfaction Rating",
      y="Minerals_Scaled",
      title="Scatter plot of Value vs Minerals_Scaled"
    ) +
    theme_minimal() +
    theme(plot.title=element_text(hjust=0.5)) +
    geom_hline(yintercept=seq(0, 10, by=1), linetype="dotted", color="gray")
```

```{r}
exports <- df_mr %>%
  filter(Flow.name=="Exports") %>%
  select(Country, Category, Flow.name, "X2023") %>%
  mutate(X2023 = X2023 / 1e6)

top_countries <-exports %>%
                group_by(Country) %>%
                summarise(total=sum(X2023, na.rm=TRUE)) %>%
                slice_max(order_by=total, n=5) %>%
                pull(Country)

df_top5 <- exports %>%
           filter(Country %in% top_countries)

pivot_df <- df_top5 %>%
            pivot_wider(
              names_from = Category,
              values_from = X2023,
              values_fill = 0
            )

print(pivot_df)
pivot_long <- pivot_df %>%
              pivot_longer(
                cols=-c(Country, Flow.name),
                names_to="Category",
                values_to="Value"
              )

ggplot(pivot_long, aes(x=Country, y=Value, fill=Category)) +
      geom_bar(stat="identity", position="stack", color="black", linewidth=0.5) +
      scale_fill_viridis_d() +
      labs(
        title="Top 5 Countries by Total Exports in 2023 (Mt)",
        x="Country",
        y="Export Value",
        fill="Export Category"
      ) +
      theme_minimal(base_size=14) +
      theme(
        legend.position="right",
        legend.title=element_text(size=10),
        axis.text.x = element_text(angle=45, hjust=1)
      )
```

```{r}
normalize <- function(x) {
  (x - min(x, na.rm=TRUE)) / (max(x, na.rm=TRUE) - min(x, na.rm=TRUE))
}

df_current <- filter_wellbeing(df_wb, measure="Life satisfaction", domain="Subjective well-being", years=c(2021)) %>%
  rename(c(Country="Reference.area", Year="TIME_PERIOD", Value="OBS_VALUE"))

merged_df <- df_current %>%
    inner_join(annual_earnings, by="Country") %>%
    select(all_of("Country", Value_x=Value.x, Value_y=Value.y))
```

``` r
```

```         
```
